---
title: "nfl_databowl_pt1"
output: html_document
date: "2025-09-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages

```{r, warning=F, message=f}
### Installing packages
install.packages("nflreadr")
install.packages("tidyverse")
install.packages("lubridate")
install.packages("janitor")
install.packages("scales")
install.packages("gt")
install.packages("ggrepel")
install.packages("zoo")
```

```{r}
### Loading libraries
library(nflreadr)    # data access
library(tidyverse)   # dplyr/ggplot/readr, etc.
library(lubridate)   # dates/times
library(janitor)     # clean_names, tabyl
library(scales)      # pretty labels
library(gt)          # presentation tables (not saved here)
library(ggrepel)     # labels for charts
library(zoo)         # rolling calcs
```

# 1.1 Load and Prepare Data

```{r}
### Loading Colt's Game 1 Data:

# ---- Parameters ----
team   <- "IND"
season <- 2025

# ---- 1) Identify Colts' first regular-season game in 2025 ----
sched <- load_schedules(seasons = season) %>% clean_names()

colts_g1 <- sched %>%
  filter(game_type == "REG",
         season == !!season,
         home_team == !!team | away_team == !!team) %>%
  arrange(gameday, week) %>%
  slice(1)

stopifnot(nrow(colts_g1) == 1)

game_id <- colts_g1$game_id
week    <- colts_g1$week
gameday <- colts_g1$gameday
home_tm <- colts_g1$home_team
away_tm <- colts_g1$away_team
opp     <- ifelse(home_tm == team, away_tm, home_tm)

message(glue::glue("Found Colts Week {week} ({gameday}): {away_tm} @ {home_tm} | game_id = {game_id}"))

# ---- 2) Pull core data for this game ----
pbp_g1 <- load_pbp(seasons = season) %>%
  clean_names() %>%
  filter(game_id == !!game_id)

rw_g1 <- load_rosters_weekly(seasons = season) %>%
  clean_names() %>%
  filter(week == !!week, team %in% c(team, opp))

officials_g1 <- load_officials(seasons = season) %>%
  clean_names() %>%
  filter(game_id == !!game_id)

players_master <- load_players() %>% clean_names()
```

1. The load_schedules() function loads the schedule data. We specified a specific season to the arguments.

2. We can run the filter() function with the arguments that game_type == ‘REG’ and then we filtered out the games that did not contain the colts (home_team or away_team != ‘IND’). Next, we arranged the data by gameday and week, and then only selected the 1st week by using the slice(1) function.

3. This ensures that the colts schedule is in order from week 1 first before slicing just the 1st row. Thus, we actually properly select the 1st game.

# 1.2 Data Integrity Checks

```{r}
# ---- A) Structural sanity: dimensions & basic types ----
cat("\n--- STRUCTURE ---\n")
cat("PBP rows/cols:", nrow(pbp_g1), ncol(pbp_g1), "\n")
cat("Weekly roster rows/cols:", nrow(rw_g1), ncol(rw_g1), "\n")
cat("Officials rows/cols:", nrow(officials_g1), ncol(officials_g1), "\n\n")
```

```{r}
# ---- B) Key constraints & duplicates ----
# Expect play-level key uniqueness by (game_id, play_id)
dup_plays <- pbp_g1 %>%
  count(game_id, play_id) %>%
  filter(n > 1)

if (nrow(dup_plays) > 0) {
  warning("Duplicate plays detected in (game_id, play_id). Inspect `dup_plays`.")
} else {
  cat("Key check passed: (game_id, play_id) appear unique for this game.\n")
}

# For weekly rosters, (team, gsis_id) often repeats by position slots; check basic duplication
dup_roster <- rw_g1 %>%
  count(team, gsis_id, week) %>%
  filter(n > 1)
if (nrow(dup_roster) > 0) {
  message("Note: roster has repeated (team, gsis_id, week) rows—can be normal when multiple rows exist per role/slot.")
}
```

1. nrow(dup_plays) tells us how many rows in the dup_plays table there are. The dup_plays table is created to ensure that no plays were duplicated in the dataset by only including pairs of game_id and play_id that appear more than once (duplicates). We would want this dataset to be empty (nrow = 0).


```{r}
# ---- C) Referential integrity (players present in master) ----
# Attempt to match weekly roster players to master by gsis_id (best available common key)
if ("gsis_id" %in% names(rw_g1) && "gsis_id" %in% names(players_master)) {
  roster_master_match <- rw_g1 %>%
    mutate(in_master = gsis_id %in% players_master$gsis_id) %>%
    summarize(match_rate = mean(in_master, na.rm = TRUE))
  cat("Player master match rate (weekly roster -> master by gsis_id):",
      round(roster_master_match$match_rate*100, 1), "%\n")
} else {
  message("gsis_id not present to check roster->player master referential integrity.")
}
```

```{r}
# ---- D) Missingness (critical analytical fields) ----
na_rate <- function(x) mean(is.na(x))
na_summary <- pbp_g1 %>%
  summarize(
    na_epa          = na_rate(epa),
    na_success      = na_rate(success),
    na_posteam      = na_rate(posteam),
    na_defteam      = na_rate(defteam),
    na_yardline_100 = na_rate(yardline_100),
    na_down         = na_rate(down),
    na_ydstogo      = na_rate(ydstogo),
    na_play_type    = na_rate(play_type)
  )
cat("\n--- MISSINGNESS (PBP) ---\n"); print(na_summary)
```

```{r}
# ---- E) Range & logical checks ----
# yardline_100 should be within [0, 100]
bad_yardline <- pbp_g1 %>% filter(!is.na(yardline_100) & (yardline_100 < 0 | yardline_100 > 100))
if (nrow(bad_yardline) > 0) warning("Out-of-range yardline_100 values observed.")

# down should be 1–4 (ignore special plays like kickoffs which may have NA)
bad_down <- pbp_g1 %>% filter(!is.na(down) & !(down %in% 1:4))
if (nrow(bad_down) > 0) warning("Unexpected down values outside 1–4.")

# ydstogo should be positive for standard plays
bad_ydstogo <- pbp_g1 %>% filter(!is.na(ydstogo) & ydstogo <= 0)
if (nrow(bad_ydstogo) > 0) message("Non-positive ydstogo rows exist (could be goal-to-go anomalies or data glitches).")

# game_id consistency
if (length(unique(pbp_g1$game_id)) != 1) warning("Multiple game_ids in pbp_g1; filter may be off.")

# posteam/defteam should be one of the two teams (allow NA on non-plays)
valid_teams <- c(team, opp)
weird_teams <- pbp_g1 %>%
  filter(!is.na(posteam) & !(posteam %in% valid_teams)) %>%
  distinct(posteam)
if (nrow(weird_teams) > 0) warning("Found posteam not matching Colts or opponent: check `weird_teams`.")
```

2. Checking yardline_100 range ensures that no play is run outside of the 100 yards of play. Yardline_100 determines what yard line on the field the play starts on (100 is the own endzone, 0 is the opponents endzone), so any plays outside of that range would be improperly stored. 

3. Rows with ydstogo <= 0 appear to represent breaks in play (start/end of game/quarter and timeouts) as well as extra point kicks/2 point conversions and kickoffs. These could also be issues in data entry.

```{r}
# ---- F) Categorical consistency (team abbreviations) ----
cat("\n--- TEAM ABBREVS IN PBP ---\n")
print(sort(unique(na.omit(pbp_g1$posteam))))
print(sort(unique(na.omit(pbp_g1$defteam))))
```

# 2.1: Create points_scored

```{r}
pbp_g1 <- pbp_g1 %>%
  mutate(
    points_scored = case_when(
      touchdown == 1 ~ 6,                               # Touchdowns
      field_goal_result == "made" ~ 3,                  # Field goals
      safety == 1 ~ 2,                                  # Safeties
      extra_point_result == "good" ~ 1,                 # PAT kicks
      two_point_attempt == 1 & two_point_conv_result == "success" ~ 2, # 2PT conversions
      TRUE ~ 0
    )
  )
```

1. case_when() allows us to specifically determine how many points to assign to the points_scored variable based on if the play was a scoring play. There are variables related to each type of scoring play that indicate if points of that type were scored on that play (e.g. touchdown == 1 means a touchdown was scored), so we can assign the proper amount of points to each row/play based on that.

2. If we did not include two-point conversions, any plays where a two point conversion was successful would not be properly accounted for. We would then improperly record the scoring if there were successful 2 point conversions. This could affect many other metrics in the end.

# 2.2 Quarter by Quarter Scoring

```{r}
score_by_q <- pbp_g1 %>%
  filter(!is.na(qtr), qtr %in% 1:4) %>%
  group_by(qtr) %>%
  summarize(
    ind_points = sum(if_else(posteam == "IND", points_scored, 0), na.rm = TRUE),
    opp_points = sum(if_else(posteam == opp,   points_scored, 0), na.rm = TRUE),
    plays = n(),
    .groups = "drop"
  )
cat("\n--- SCORE BY QUARTER (derived from PBP points events) ---\n")
print(score_by_q)
```

1. Grouping by qtr before summarize is essential because it ensures that we summarize the points scored by each quarter rather than over the entire game.

2. To calculate 1st vs 2nd half points, we could group by game_half instead of grouping by qtr.

# 2.3: Offensive Play Mix & Efficiency

```{r}
# ---- 4) Offensive profile: play mix & efficiency ----
### Filtering out NA rows in posteam to start
pbp_g1_filtered <- pbp_g1 %>%
  filter(!is.na(posteam))

off_mix <- pbp_g1_filtered %>%
  filter(!is.na(posteam)) %>%
  mutate(play_family = case_when(
    play_type %in% c("run") ~ "Run",
    play_type %in% c("pass") ~ "Pass",
    play_type %in% c("qb_kneel","qb_spike") ~ "Clock",
    play_type %in% c("no_play","timeout") ~ "Other",
    TRUE ~ "Other"
  )) %>%
  group_by(posteam, play_family) %>%
  summarize(
    plays          = n(),
    share          = n()/nrow(pbp_g1_filtered),
    epa_per_play   = mean(epa, na.rm = TRUE),
    success_rate   = mean(success, na.rm = TRUE),
    yards_per_play = mean(yards_gained, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(posteam, desc(plays))
cat("\n--- OFFENSIVE MIX & EFFICIENCY ---\n"); print(off_mix, n = 50)
```

1. We create a play_family variable instead of just using play_type because we wanted to group the plays differently than just how play_type had them listed. QB kneels and Qb spikes were combined into the “Clock” play family because they are done to manage the clock and aren’t plays designed to gain yards or score. No_play and timeout were also combined because both don’t result in a play being run.

2. To verify that the share of plays adds up to 1, we can run sum(off_mix$share) and verify that this = 1. In this case, after filtering out the rows with N/A in posteam before completing the average calculation in the creation of share, we ensure that it adds up to 1.

```{r}
sum(off_mix$share)
```

# 2.4 Situational Analysis

```{r}
# ---- 5) Situational analysis ----
situ <- pbp_g1 %>%
  filter(!is.na(down), !is.na(ydstogo), play_type %in% c("pass","run")) %>%
  mutate(
    ytg_bucket = case_when(
      ydstogo <= 2 ~ "1-2",
      ydstogo <= 5 ~ "3-5",
      ydstogo <= 10 ~ "6-10",
      TRUE ~ "11+"
    ),
    down_lab = paste0("Down ", down)
  ) %>%
  group_by(posteam, down_lab, ytg_bucket) %>%
  summarize(
    plays = n(),
    success_rate = mean(success, na.rm = TRUE),
    epa_per_play = mean(epa, na.rm = TRUE),
    .groups = "drop"
  )
ggplot(situ %>% filter(posteam %in% c("IND", opp)),
       aes(x = ytg_bucket, y = success_rate, group = posteam, color = posteam)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~down_lab) +
  scale_y_continuous(labels = percent) +
  labs(title = "Success rate by down & yards-to-go",
       x = "Yards to go (bucket)", y = "Success rate", color = "Offense") +
  theme_minimal()
```

1. case_when() helps us determine which situations in terms of yards to go should be placed in each bucket. Anything <= 2 is put in 1-2 yards left, then the remaining that have <= 5 are in 3-5, etc. while creating the ytg_bucket.

2. For 1st down, field position didn't really seem to make a huge difference for success rate, as it bounced around between 0 to 100% for the Colts, and was only shown at ~40% for the Dolphins for 6-10 yards to go. On 2nd down, there wasn't a ton of variance between the yard buckets, but success was higher for both teams with either 1-2 yards to go or 11+ yards to go, with the lowest at 3-5 yards for each. For 3rd down, both teams had the highest success at 1-2 yards, and it decreased as yards to go increased.

# 2.5: Field Position and Drive Analysis

```{r}
# Field position zones
fieldpos <- pbp_g1 %>%
  filter(!is.na(yardline_100), play_type %in% c("run","pass")) %>%
  mutate(
    field_zone = case_when(
      yardline_100 >= 80 ~ "Own 20-0",
      yardline_100 >= 50 ~ "Own 49-21",
      yardline_100 >= 21 ~ "Opp 49-21",
      TRUE ~ "Red Zone (<=20)"
    )
  ) %>%
  group_by(posteam, field_zone) %>%
  summarize(
    plays = n(),
    epa_per_play = mean(epa, na.rm = TRUE),
    success_rate = mean(success, na.rm = TRUE),
    .groups = "drop"
  )
ggplot(fieldpos %>% filter(posteam %in% c("IND", opp)),
       aes(x = field_zone, y = epa_per_play, fill = posteam)) +
  geom_col(position = position_dodge(width = 0.8)) +
  coord_flip() +
  labs(title = "EPA/play by field zone",
       x = "", y = "EPA per play", fill = "Offense") +
  theme_minimal()
```

```{r}
# ---- 6) Drive lens ----
drive_sum <- pbp_g1 %>%
  filter(!is.na(drive), !is.na(posteam)) %>%
  group_by(posteam, drive) %>%
  summarize(
    plays        = n(),
    yards        = sum(yards_gained, na.rm = TRUE),
    points       = sum(points_scored, na.rm = TRUE),   # <-- FIXED
    epa          = sum(epa, na.rm = TRUE),
    success_rate = mean(success, na.rm = TRUE),
    ended_score  = any(touchdown == 1 | field_goal_result == "made" | safety == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(posteam, desc(points), desc(yards))

print(drive_sum)
```

1. drive might be missing or some plays because they could be the kickoffs, extra point plays, or start/end of game/half scenarios where a drive isn’t really occurring.

2. A negative total EPA on a drive means that the drive was very inefficient and the team would definitely not be expected to score any points on that drive. These occurred on drives where the team did not score, but theoretically could be positive even if points weren’t scored if there was a punt or the end of a half or something like this.

# 2.6 Player Involvement

```{r}
colts_targets <- pbp_g1 %>%
  filter(posteam == team, pass == 1) %>%
  count(receiver_player_name, sort = TRUE, name = "targets")

colts_rushers <- pbp_g1 %>%
  filter(posteam == team, rush == 1) %>%
  count(rusher_player_name, sort = TRUE, name = "rush_att")

epa_by_receiver <- pbp_g1 %>%
  filter(posteam == team, pass == 1, !is.na(receiver_player_name)) %>%
  group_by(receiver_player_name) %>%
  summarize(
    targets = n(),
    epa_sum = sum(epa, na.rm = TRUE),
    epa_mean = mean(epa, na.rm = TRUE),
    .groups = "drop"
  ) %>% arrange(desc(epa_sum))

ggplot(epa_by_receiver, aes(x = targets, y = epa_sum, label = receiver_player_name)) +
  geom_point(size = 3) +
  geom_text_repel(min.segment.length = 0) +
  labs(title = "Colts receivers: total EPA vs targets",
       x = "Targets", y = "Total EPA") +
  theme_minimal()
```

1. We filter for pass = 1 when counting targets because pass is an indicator variable on if a passing play occurs, so there would only be an intended target on these plays.

```{r}
print(epa_by_receiver)
```

2. Mo Alie-Cox had the highest EPA/touch, as he had an average EPA per touch of 1.361

# 2.7: Penalties & Special Teams

```{r}
penalties <- pbp_g1 %>%
  filter(!is.na(penalty)) %>%
  mutate(off_def = if_else(penalty_team == posteam, "Offense", "Defense")) %>%
  count(penalty_team, off_def, penalty_type, sort = TRUE)
cat("\n--- PENALTIES ---\n"); print(penalties, n = 30)

st_fg <- pbp_g1 %>%
  filter(play_type == "field_goal") %>%
  count(posteam, field_goal_result)
st_punt <- pbp_g1 %>%
  filter(play_type == "punt") %>%
  summarize(punts = n(), avg_punt_yards = mean(yards_gained, na.rm = TRUE))
cat("\n--- SPECIAL TEAMS ---\n"); print(list(field_goals = st_fg, punts = st_punt))
```

```{r}
total_penalties <- penalties %>%
  group_by(penalty_team) %>%
  summarize(
    total_penalties = sum(n)
  )

print(total_penalties)
```

1. Indianapolis committed more penalties (4) than Miami (3).

2. A key penalty, such as one committed by the defensive team on 3rd down that gives the offense a 1st down, can significantly swing the EPA in the offense's favor. Thus, the win probability would increase for the offense because they are expected to score more points.

# 2.8: Time-Based Trends

```{r}
# ---- 9) Time-based trends ----
epa_by_q <- pbp_g1 %>%
  filter(qtr %in% 1:4, play_type %in% c("pass","run")) %>%
  group_by(posteam, qtr) %>%
  summarize(
    epa_per_play = mean(epa, na.rm = TRUE),
    success_rate = mean(success, na.rm = TRUE),
    .groups = "drop"
  )
ggplot(epa_by_q, aes(x = factor(qtr), y = epa_per_play, group = posteam, color = posteam)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  labs(title = "EPA/play by quarter", x = "Quarter", y = "EPA/play", color = "Offense") +
  theme_minimal()

# Rolling pass rate (10-play window) for Colts
colts_seq <- pbp_g1 %>%
  filter(posteam == team, play_type %in% c("pass","run")) %>%
  mutate(play_index = row_number(),
         pass_flag  = if_else(play_type == "pass", 1, 0)) %>%
  arrange(play_index)
if (nrow(colts_seq) >= 10) {
  colts_seq_roll <- colts_seq %>%
    mutate(pass_rate_10 = zoo::rollapply(pass_flag, width = 10, FUN = mean,
                                         align = "right", fill = NA_real_))
  ggplot(colts_seq_roll, aes(x = play_index, y = pass_rate_10)) +
    geom_line(linewidth = 1) +
    scale_y_continuous(labels = percent) +
    labs(title = "Colts rolling pass rate (10-play window)",
         x = "Offensive play index", y = "Pass rate") +
    theme_minimal()
}
```

1. For the Colts, the rolling pass rate decreased after halftime because they were winning, and thus wanted to run down the clock. Running plays are less likely to result in stoppages of the clock due to incompletions or running out of bounds than passing plays, so the Colts would run more.

2. A spike or dip in rolling pass rate would indicate a shift in the game context. This could be due to clock management, injuries, turnovers, or score changes.







